Linear Stochastic Approximation- Constant Step-Size and Iterate Averaging

### 2017-05-14 related.tex
recent updates

\paragraph{General LSA:} Analysis of LSA with CS-PR goes back to the work by \citet{polyak-judisky}, wherein they considered the additive noise setting i.e., $A_t=A$ for some deterministic Hurwitz matrix $A\in \R^{\dcd}$. A key improvement in this paper is that we consider the `multiplicative' noise case, i.e., $A_t$ is a random matrix.
\paragraph{SGD for Linear Regression} As mentioned in \Cref{sec:intro}, we were motivated by the ``remarkable'' results obtained by \citet{bach}, where the authors analyse the performance of SGD for the problem of linear regression with quadratic loss.
Here, $(x_t,y_t)\in \R^d\times \R,\,t\geq 1$ is an $i.i.d$ a bounded (by $B$) data sequence such that $\EE{\norm{x_t}^2}$ and $\EE{y_t^2}$ are finite. $x_t,\,t\geq 1$ are the features and $y_t$ are real labels and the linear regression problem constitutes minimizing the quadratic loss function $f(\theta)=\EE{\left(\ip{x_t,\theta}-y_t\right)^2}$. SGD with CS-PR to minimize $f(\theta)$ can be given as below:
\begin{subequations}\label{eq:linreg}
\begin{align}
\theta_t&=\theta_{t-1}-\alpha x_t\left(\ip{x_t,\theta_{t-1}} -y_t\right)\\
\thh_t&=\frac{1}{t+1}\sum_{s=0}^t \theta_t
\end{align}
\end{subequations}
Here $\nabla f(\theta_{t-1})=x_t\left(\ip{x_t,\theta_{t-1} -y_t}\right)$ is the \emph{stochastic gradient} of the objective $f(\theta)$. It follows from the $i.i.d$ assumption on $(x_t,y_t)$ that $A_t=x_t x_t^\top$ and $b_t=y_t x_t$ are also $i.i.d$, and with this identification one can see that \eqref{linreg} is of the form in \eqref{conststep}.
%In this case $A_P$ is SPD.
In the linear regression setting, the prediction error is given by $\EE{\norm{\theta-\ts}^2_{A_P}}$.
%$(ii)$ the MSE in \eqref{eq:mselinreg} is measured with respect to $A_P$, $(iii)$ the noise $%\epsilon_t=y_t-\ip{\ts,x_{t-1}}$ is structured i.e., $\EE{\epsilon_t^2 x_t x_t^\top}\leq R A_P$ for some constant $R>0$.
The problem class considered by them belongs to $\P_{\text{PSD},B}$ (see \Cref{def:pclass} ) and hence is weakly admissible with a uniform step-size choice. In order to produce a rate of $\frac{C}{t}$ (where $C>0$ is a problem independent constant) for the bias term \citet{bach} do not rely on admissibility, but instead use the fact that the MSE $\EE{\norm{\thh_t-\ts}^2_{A_P}}$ for the linear prediction problem is measured in the general quadratic norm induced by $A_P$ and the fact that the noise is `structured' i.e., the noise is bounded by $R\,A_P$ for some constant $R>0$. At the same time, the bias term decays at a rate $\frac{C'}{t^2}$, where the constant $C'$ contains the factor $(\theta_0-\ts)^\top A^{-1}_P (\theta_0-\ts)$ and is dependent on the condition number.

@@

\end{tabular}
}
\label{tb:rates}
\end{table}
\end{comment}
Implications
\begin{itemize}
\item It follows that for each instance of blah blah
\item Bias vairance trade off
\end{itemize}
\begin{comment}
\paragraph{Related Work in RL} The various temporal difference learning algorithms can be written as in \Cref{table:td}. The TD(0) algorithm is an on-policy algorithm ($\mu_t=1$) and is unstable in the general {off-policy} setting. The gradient temporal difference (GTD) algorithm \cite{gtd,gtd2,gtdmp} is stable in off-policy setting as well. Recent work by \cite{korda-prashanth} derive finite time bounds for the TD(0) for two different cases of step-sizes one without PR averaging and other with PR averaging. In the case without PR averaging, \cite{korda-prashanth} consider step-sizes of the form $\frac{c_0c}{c+t}$, where $c_0$ and $c$ are problem dependent positive constants, to dervie a rate of $O(\frac{1}{t})$ for the MSE.  In the case with PR averaging (row $1$ of \Cref{tb:rates}), \cite{korda-prashanth} consider step-size of the from $c_0\left(\frac{c}{c+t}\right)^\delta,\,\delta \in (1/2,1)$ where $c_0, c$ are problem independent positive constants, and show that this choice yeilds a rate of $O(\frac{1}{t})$ for $\delta \ra 1$ and $t\geq t_0$ (where $t_0$ is a problem dependent time instant). A possible limiting feature of the results by \cite{korda-prashanth} is that either the problem knowledge is required or the bounds hold only after a problem dependent time. A centered TD(0) variant is also considered by \cite{korda-prashanth}, however, a projection operator is required to keep the iterates bounded. Also, the bounds by \cite{korda-prashanth} don't have an explicit bias variance decomposition. The works on GTD and TDC by \cite{gtd,gtd2} show only asymptotic convergence w.p.1. Recently, the work by \cite{gtdmp} shows a rate of $\frac{1}{\sqrt{t}}$ for GTD with use of projection operation, and also proposed a variant namely GTD-Mirror-prox based on stochastic mirror-prox ideas which achieves $\frac{1}{\sqrt{t}}$ rate without use of projection. In comparison, we don't use of projections, our bounds hold for all time $t$, and the bounds are dependent on the problem instance only to the extent of choosing a step-size that leads to stable behavior, in which case we show that $O\frac{1}{t}$ rate holds (the scaling factor hidden in the $O$ expression is problem dependent). Further, our results provide explicit bias variance decomposition.
\end{comment}
An interesting fact to note is while theoretical analysis for GTD/GTD2/TDC are presented for the case of diminishing step-size, the experimental results are reported for constant step-size choices \cite{gtd2,gtdmp}.

>>>>

\paragraph{SGD for Linear Regression} As mentioned in \Cref{sec:intro}, we were motivated by the ``remarkable'' results obtained by \citet{bach}, where the authors analyse the performance of SGD for the problem of linear regression with quadratic loss. The problem class considered by them belongs to $\P_{\text{PSD},B}$ (see \Cref{def:pclass} ) and hence is weakly admissible with a uniform step-size choice. In order to produce a rate of $\frac{C}{t}$ (where $C>0$ is a problem independent constant) for the bias term \citet{bach} do not rely on admissibility, but instead use the fact that the MSE $\EE{\norm{\thh_t-\ts}^2_{A_P}}$ for the linear prediction problem is measured in the general quadratic norm induced by $A_P$ and the fact that the noise is `structured' i.e., the noise is bounded by $R\,A_P$ for some constant $R>0$. At the same time, the bias term decays at a rate $\frac{C'}{t^2}$, where the constant $C'$ contains the factor $(\theta_0-\ts)^\top A^{-1}_P (\theta_0-\ts)$ and is dependent on the condition number.
\paragraph{General LSA:} Analysis of LSA with CS-PR goes back to the work by \citet{polyak-judisky}, wherein they considered the additive noise setting i.e., $A_t=A$ for some deterministic Hurwitz matrix $A\in \R^{\dcd}$.
\paragraph{Our Results:} A key improvement in this paper is that we consider the `multiplicative' noise case, i.e., $A_t$ is a random matrix. To tackle the multiplicative noise we use newer analysis introduced by \citet{bach}. However, since the general LSA setting does not enjoy special structures of the SGD setting, we make use of Jordan decomposition and similarity transformations in a critical way to prove our results.

@@

The \emph{temporal difference} learning algorithm or simply known as TD(0), is the most basic of the class of TD algorithms. An important shortcoming of TD(0) was its instability in the \emph{off-policy} case, and was successfully mitigated by the \emph{gradient temporal difference} learing GTD \cite{gtd2}. The GTD was first proposed by \citet{gtd} and its variants namely GTD2 and TDC were proposed later by \citet{gtd2}. However, the initial convergence analysis for GTD/GTD2/TDC were only asymptotic in nature \cite{gtd,gtd2}.\par
The most relevant to our results are those by \citet{korda-prashanth} in TD(0) and by \citet{gtdmp} in GTD. For the TD(0) case,  diminishing step-sizes $alpha_t=O(\frac{1}{t})^\beta,\beta \in(0,1)$ with PR averaging is showed to exhibit a rate of $O(\frac{1}{t})$ decay for the MSE \cite{korda-prashanth}. In the case of GTD/GTD2 diminishing step-sizes $\alpha_t=O(\frac{1}{\sqrt{t}})$, projection of iterates and PR-averaging leads to a rate of $O(\frac{1}{\sqrt{t}})$ is achieved for the prediction error $\norm{A_P\thh_t-b_P}^2$ with high probability. \citet{gtdmp} also suggests a new version of GTD, based on stochastic mirror prox ideas, called the GTD-Mirror-Prox which also achieves $O(\frac{1}{\sqrt{t}})$ rate for $\norm{\eh_t}$ with high probability under similar step-size choice that was used by them for the GTD.
\paragraph{Our Results} \Cref{th:rate} applies to TD(0)/TDC/GTD/GTD2/GTDMP with CS-PR. Finite time performance of TDC was unkown in prior literature and our work provides a first step by understanding TDC with CS-PR. Further, our results show that the error in the GTD class of algorithms decay at $O(\frac{1}{t})$ rate instead of $O(\frac{1}{\sqrt{t}})$, (without use of projection or mirror maps) which is an improvement. In comparison to the TD(0) results by \citet{korda-prashanth}, \Cref{th:rate} is better in that it provides the bias/variance decomposition. However, it is important to note that \citet{korda-prashanth} handle the Markov noise case which is not dealt with in this paper.
\paragraph{Related Work in RL} The various temporal difference learning algorithms can be written as in \Cref{table:td}. The TD(0) algorithm is an on-policy algorithm ($\mu_t=1$) and is unstable in the general {off-policy} setting. The gradient temporal difference (GTD) algorithm \cite{gtd,gtd2,gtdmp} is stable in off-policy setting as well. Recent work by \cite{korda-prashanth} derive finite time bounds for the TD(0) for two different cases of step-sizes one without PR averaging and other with PR averaging. In the case without PR averaging, \cite{korda-prashanth} consider step-sizes of the form $\frac{c_0c}{c+t}$, where $c_0$ and $c$ are problem dependent positive constants, to dervie a rate of $O(\frac{1}{t})$ for the MSE.  In the case with PR averaging (row $1$ of \Cref{tb:rates}), \cite{korda-prashanth} consider step-size of the from $c_0\left(\frac{c}{c+t}\right)^\delta,\,\delta \in (1/2,1)$ where $c_0, c$ are problem independent positive constants, and show that this choice yeilds a rate of $O(\frac{1}{t})$ for $\delta \ra 1$ and $t\geq t_0$ (where $t_0$ is a problem dependent time instant). A possible limiting feature of the results by \cite{korda-prashanth} is that either the problem knowledge is required or the bounds hold only after a problem dependent time. A centered TD(0) variant is also considered by \cite{korda-prashanth}, however, a projection operator is required to keep the iterates bounded. Also, the bounds by \cite{korda-prashanth} don't have an explicit bias variance decomposition. The works on GTD and TDC by \cite{gtd,gtd2} show only asymptotic convergence w.p.1. Recently, the work by \cite{gtdmp} shows a rate of $\frac{1}{\sqrt{t}}$ for GTD with use of projection operation, and also proposed a variant namely GTD-Mirror-prox based on stochastic mirror-prox ideas which achieves $\frac{1}{\sqrt{t}}$ rate without use of projection. In comparison, we don't use of projections, our bounds hold for all time $t$, and the bounds are dependent on the problem instance only to the extent of choosing a step-size that leads to stable behavior, in which case we show that $O\frac{1}{t}$ rate holds (the scaling factor hidden in the $O$ expression is problem dependent). Further, our results provide explicit bias variance decomposition.

### 2017-05-14 stepsizes.tex
recent updates

\section{Automatic Tuning of Step-Sizes}

@@

\caption*{Automatic Tuning of Constant Step-Size}

@@

\STATE{Initialize: $\theta_0$, $\alpha=1$, $i=1$, $\rho_0=0$}

@@

\STATE{$\theta_t=\theta_{t-1}+\alpha (b_t-A_t\theta_{t-1})$}
\IF{$t (mod k)==0$}
\STATE{$\rho_{i}=\norm{\theta_t}$}
\IF{IsUnstable($\{\rho_j,j=1,\ldots,i\}$)==1}

@@

\STATE{$i=i+1$}

@@

@@

In \Cref{alg:tuning} IsUnstable$(\rho_j,j=1,\ldots,i\}$) is a routine whose input is $\norm{\theta_t}$ measured at time instances $k$ apart. It is easy to show from Gronwall type arguments that the $\norm{\theta_t}\leq (1-exp{\beta t})$, where $\beta>0$ implies stability and $\beta<0$ imples instability. From this Gronwall based inequality, we observe that when the iterates are stable the ratio $\frac{\rho_{k+1}}{\rho_k},k=1,\ldots,i-1$ should be `roughly' decreasing (because there is the iterates are persistently corrupted by noise) and should converge to $1$. This condition on the sequence $\frac{\rho_{k+1}}{\rho_k}$ can be checked in practice with appropriate tolerances.

>>>>

\section{Automatic Tuning of Step-Sizes}
It is straightforwatd to see from \eqref{eq:lsaintro} that $\alpha_t$ cannot be asymptotically increasing. We now present some heuristic arguments in favour of a constant step-size over asymptotically diminishing step-sizes in \eqref{eq:lsaintro}.
It has been observed that when the step-sizes of form $\alpha_t=\frac{c}{t}$ or $\alpha_t=\frac{c}{c+t}$ (for some $c>0$) are used, the MSE, $\EE{\norm{\theta_t-\ts}^2}$, is not robust to the choice of $c>0$ \cite{korda-prashanth,bach-moulines}. In particular only a $O(\frac{1}{t^{\mu c}/2})$ decay can be achieved for the MSE, where $\mu$ is the smallest positive part of the $A_P$ \cite{bach-moulines}. Note that, in the case of LSA with CS-PR, \Cref{th:rate} guarantees a $O(\frac{1}{t})$ rate of decay for the MSE and the problem dependent quantities affect only the constants and not the exponent. Also, in the case of important TD algorithms such as GTD/GTD2/TDC, while the theoretical analysis uses diminishing step-sizes, the experimental results are with a constant step-size or with CS and PR averaging \cite{gtd2,gtdmp}. Independently, \citet{dann} also observe in their experiments that a constant step-size is better than diminishing step-sizes.\par
Motivated by \Cref{th:rate} and also by the usage of constant step-size in practice \cite{dann,gtd2,gtdmp}, we suggest a heuristic algorithm to tune the constant step-size (see \Cref{alg:tuning}).

@@

\caption{Automatic Tuning of Constant Step-Size}

@@

\STATE{Initialize: $\theta_0$, $\alpha=\alpha_{\max}$, $k$, $T$ }

@@

\STATE{$\theta_t=\theta_{t-1}+\alpha (b_t-A_t\theta_{t-1}), \thh_t=\thh_{t-1}+\frac{1}{t+1}(\theta_t-\thh_{t-1})$}
\IF{$IsUnstable(\norm{\theta_t},\ldots,\norm{\theta_{(t-kT)\wedge 0}})=1$}

@@

@@

In \Cref{alg:tuning} $T>0$ is a time epoch and $k$ is a given integer and $\alpha_{\max}>0$ is the maximum step-size that is allowable. It is easy to show using a Gronwall argument that in \Cref{alg:tuning} $\norm{\theta_t}\leq (1-e^{\beta t})$, where ($\beta>0)$) $\beta<0$ implies (in)stability of iterates. Using this fact, we observe that the sequence $r_i=\frac{\norm{\theta_{(t-kT+iT)\wedge 0}}}{\norm{\theta_{(t-kT+(i-1)T)\wedge 0}}},i=1,\ldots,k$ should be `roughly' (making allowance for the presistent noise) decreasing and converge to $1$. The $IsUnstable()$ routine in \Cref{alg:tuning} can take this sequence of $r_i$s as input, and in the case of instability, the step-size can be scaled down by a factor of $2$ (or any constant factor).

@@

In \Cref{alg:tuning} IsUnstable$(\rho_j,j=1,\ldots,i\}$) is a routine whose input is $\norm{\theta_t}$ measured at time instances $k$ apart.
It is easy to show from Gronwall type arguments that the $\norm{\theta_t}\leq (1-exp{\beta t})$, where $\beta>0$ implies stability and $\beta<0$ imples instability. From this Gronwall based inequality, we observe that when the iterates are stable the ratio $\frac{\rho_{k+1}}{\rho_k},k=1,\ldots,i-1$ should be `roughly' decreasing (because there is the iterates are persistently corrupted by noise) and should converge to $1$. This condition on the sequence $\frac{\rho_{k+1}}{\rho_k}$ can be checked in practice with appropriate tolerances.
